# Manipulate HTML tags from R with htmltools

**Learning objectives:**

- Generate HTML elements from R.
- Modify the attributes of HTML
elements from R.
- Access chidren elements from an
HTML element and modify them via the
`tagQuery` function.

```{r, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

## Writing HTML tags from R

- Install the **htmltools** package, which allows us
to **manipulate**, **combine** and **rearrange**
HTML elements directly from R:
    ```{r}
    #| eval: false
    # CRAN
    install.packages("htmltools")
    # development version
    remotes::install_github("rstudio/htmltools")
    ```

```{r}
#| echo: false
# Render the tag instead of printing its code
library(knitr)
library(htmltools)
registerS3method(
  "knit_print", "shiny.tag",
  getS3method("print", "shiny.tag")
)
registerS3method(
  "knit_print", "shiny.tag.list",
  getS3method("print", "shiny.tag.list")
)
registerS3method(
  "knit_print", "html",
  getS3method("print", "html")
)
```

- Create an HTML element via the **htmltools** package:
    ```{r}
    library(htmltools)
    tag <- div("Hello world")
    tag
    ```

- Inside the **tag** function call, **named** elements
become attributes; and, **unnamed** elements, **children**:
    ```{r}
    div(
      id = "r-create-div",
      class = "r",
      `data-numChildren` = 1,
      div("Child")
    )
    ```

## Notations

- Some common HTML tags can be used directly without `tags$`,
such as `p`, `h1`, `div`, `span`, etc.
- Less common HTML tags require using `tags$`, such as `tags$nav`.

- Get rid of `tags$` using `withTags()`:
    ```{r}
    withTags(
      nav(
        div(),
        ul(
          li(),
          li()
        )
      )
    )
    ```

- Use `tagList()` to gather multiple HTML elements together, 
instead of `list()`,
because a special class is added (`shiny.tag.list`), which allows for extra capabilities, such as printing as HTML content:
    ```{r}
    tagList(
      p("Paragraph"),
      p("Sibling of the previous paragraph")
    )
    ```

## Adding new tags

- Create a custom tag via the `tag()` function:
    ```{r}
    customTag <- tag(
      "test",
      list(
        class = "custom",
        p("Custom tag's only child")
      )
    )
    customTag
    ```

- [Validate](https://validator.w3.org/) if the custom tag follows the HTML standards.

## Alternative way to write tags

- **Raw** HTML can be included via the `htmltools::HTML()` function.
It prevents HTML content not being properly recognized:
    ```{r}
    div("Hello <i>world</i>")

    div(HTML("Hello <i>world</i>"))
    ```

- When creating HTML elements for a Shiny app, it's recommended to 
do so via R, instead of mixing it with raw HTML:
    ```{r}
    # Same output
    HTML("<div>lorem</div>")
    div("lorem")
    ```

    ```{r}
    # Differen class
    class(HTML("<div>lorem</div>"))
    class(div("lorem"))
    ```

- The `shiny.tag` class is required for some tag-related
functions which we'll cover in the next sections.

## Playing with tags

## Tags structure {.unnumbered}

- A **shiny tag** is defined by:
    - **name**: 
        - Such as `span`, `div`, etc.
        - Accesed via `tag$name`.
    - **attributes**: 
        - Accesed via `tag$attributes`.
    - **children**: 
        - Accesed via `tag$children`.
    - **R class**: 
        - Namely `shiny.tag`.

- Example:
    ```{r}
    myTag <- div(
      class = "divClass",
      id = "first",
      h1("First child"),
      span(
        class = "child",
        id = "baby",
        "Crying"
      )
    )

    myTag
    ```

    ```{r}
    ## Access it R class
    str(myTag)

    myTag$name
    myTag$attribs
    myTag$children
    ```

  ```{r}
  # Modify the class of the second children
  myTag$
    children[[2]]$
    attribs$
    class <- "adult"

  myTag
  ```

## Practical examples {.unnumbered}

- Let's compare a common HTML structure for **select** inputs,
to what **Shiny** provides us with:
    
    ```html
    <div>
      <label id="variable-label" for="variable">Variable:</label>
      <select id="variable" class="some-class">
        <option value="cyl" selected>Cylinders</option>
        <option value="gear">Gears</option>
      </select>
    </div>
    ```

    ```{r}
    shinySelect <- shiny::selectInput(
      inputId = "selectId",
      label = "Choose",
      choices = 1:2,
      selected = 1
    )
    shinySelect
    ```

- Let's match the HTML structure for both:
    ```{r}
    # Remove class of container
    shinySelect$attribs$class <- NULL
    
    # Remove label element's class
    shinySelect$
      children[[1]]$
      attribs$class <- NULL

    # Remove extra div container for <select>
    shinySelect$children[[2]] <- 
      shinySelect$
        children[[2]]$
        children

    shinySelect
    ``` 

## Useful functions for tags {.unnumbered}

- The **htmltools** package provides us with functions
to shorten the type of HTML transformation seen in 
the previous example.

### Add attributes {.unnumbered}

- Append one or more **attributes** to an HTML
element via the `tagAppendAttributes()` function:
    ```{r}
    myTag <- div("Some div")
    myTag <- tagAppendAttributes(
      myTag,
      id = "someId",
      class = "someClass",
      `data-toggle` = "tabs"
    )
    
    myTag
    ```

### Check if an element has an specific attribute {.unnumbered}

- Function: `tagHasAttribute()`
- Example: Does this `div` element have a class?
    ```{r}
    myTag <- div(class="myClass")
    tagHasAttribute(myTag, "class")
    ```

### Get the attribute value of an element {.unnumbered}

- The `tagGetAtribute()` function returns the
target's attribute value, if it exists,
otherwise `NULL`.

- Example:
    ```{r}
    myTag <- div(class = "test", `data-toggle` = 1)
    tagGetAttribute(myTag, "class")
    tagGetAttribute(myTag, "data-toggle")
    ```

### Replace children {.unnumbered}

- The `tagSetChildren()` function creates children for
a given element, removing the already existing ones,
if they actually exist.

- Example:
    ```{r}
    myTag <- div(
      class = "parent",
      p("Son")
    )

    myTag <- tagSetChildren(
      myTag, 
      # Only daughters now
      tagList(
        p("Daughter 1"),
        p("Daughter 2")
      )
    )
    
    myTag
    ```

### Add child or children {.unnumbered}

- Add one child with `tagAppendChild()`, or many,
with `tagAppendChildren()`, to an HTML element.

- Example:
    ```{r}
    myTag <- div(
      class = "parent",
      p("Son")
    )
    myTag <- tagAppendChild(
      myTag, 
      p("Daughter")
    )

    myTag
    ```

### Build your own functions {.unnumbered}

- Create a function to remove the `nth` child:
    ```{r}
    myTag <- div(
      class = "parent", 
      p("Favourite son"),
      p("Least favourite son")
    )

    tagRemoveChild <- function(tag, n) {
      # Check if there are no children
      if (length(tag$children) == 0) {
        stop(paste(tag$name, "does not have any children!"))
      }
      tag$children[n] <- NULL
      tag
    }
    myTag <- tagRemoveChild(myTag, 2)

    myTag
    ```

- Create a function to insert the `nth` child:
    ```{r}
    tagInsertChild <- function(tag, child, position) {
      tag$children <- append(tag$children, list(child), position - 1)
      tag
    }

    res1 <- tagInsertChild(p(span("hello")), a(), 1)
    res2 <- tagInsertChild(p(span("hello")), a(), 2)

    res1
    res2
    ```

### Other functions {.unnumbered}

The **golem** package contains more functions 
to edit HTML elements, such as the 
`tagRemoveAttributes()` function.

### Conditionally set attributes {.unnumbered}

Remember that you can use the `ifelse()` function 
if you want to set some HTML element's attribute
only under *specific conditions*.

### Using the pipe operator {.unnumbered}

These last functions work well with the pipe operator:

```{r}
div(
  class = "someClass",
  h1("Hello")
) |>
  tagAppendAttributes(
    id = "someId"
  ) |>
  tagAppendChild(
    p("some extra text")
  )
```

### Programatically create children elements {.unnumbered}

- The `lapply()` and `purrr:map()` functions can be used
to programatically create children elements:
    ```{r}
    # Create 5 children
    library(purrr)
    map(1:5, \(x) span(x)) |>
      div() |>
      tagAppendAttributes(
        class = "parent"
      )
    ```

## Modern {htmltools}

- This section requires a basic understanding
of CSS selectors, therefore, let's start with a demo.

- As of **htmltools** 0.5.2, the `tagQuery`
function allows for more efficient and readable
manipulation of HTML elements, using a syntax
similar to **jQuery**.

- Two ways to modify a span which is a *third child*:
    ```{r}
    spans <- div(div(p(), map(1:5, \(i) span(i))))
    spans$children[[1]]$children[[2]][[3]]$attribs$class <- "test" 

    spans
    ```

    ```{r}
    spans <- div(div(p(), map(1:5, \(i) span(i))))
    spans <- tagQuery(spans)$
      find("span")$
      filter(\(x, i) i == 3)$
      addAttrs("class" = "amazing-tag")$
      allTags()

    spans
    ```

-  The biggest advantage is that is does not always depend
on the overall tag structure:
    ```{r}
    spans <- div(div(p(), div(map(1:5, \(i) span(i)))))
    spans <- tagQuery(spans)$
      find("span")$
      filter(\(x, i) i == 3)$
      addAttrs("class" = "amazing-tag")$
      allTags()

    spans
    ```

## Basics {.unnumbered}

- `tagQuery()` accepts an element or list of elements
as input and returns a data structure containing:
    - `$allTags()`: All elements.
    - `$selectedTags()`: Selected/transformed tags.
    Defults to `$allTags()`.

```{r}
tag_query <- tagQuery(div(p()))
class(tag_query)
tag_query
```

## Query tags {.unnumbered}

- As of now, the `tagQuery()` function only supports
simple CSS selectors ... `,`, `+`, `~ `, `data-` are not covered.

- All available query methods:

| Method | Description |
| :---: | :---: |
children	| Get all the direct descendants of each selected tag
find	| Get all descendants of each selected tag
parent	| Get the direct ancestors of each selected tag
parents	| Get all parents of each selected tag
siblings	| Get all siblings of each selected tag
filter	| Subset selected tags with CSS selectors or R function
resetSelected |	Reset set of selected tags to the root tag

- Example: Access the third tab content element
from a tabset panel with three tabs, a common case
when building custom Shiny templates.

```{r}
temp_tabs <- map(1:3, \(i) tabPanel(i, paste("Tab", i)))
tabs <- bs4Dash::tabsetPanel(.list = temp_tabs)
tabs
```

```{r}
tagQuery(tabs)$
  # div element with CSS class "tab-pane"
  find("div.tab-pane")$
  filter(\(tag, index) tagGetAttribute(tag, "data-value") == 3)$
  selectedTags()
```

## Modify tags {.unnumbered}

- The `tagQuery()` function exposes various methods
to modify **attributes** and **descendants** of the
query selection.

### Playing with attributes {.unnumbered}

| Method | Description |
| :---: | :---: |
addAttrs | Add any number of attributes to each selected tag
removeAttrs | Remove any number of attributes to each selected tag
hasAttrs | Check if the selected tag has the specified attribute(s)
addClass | Add any number of new classes to each selected tag
removeClass | Remove any number of classes to each selected tag
hasClass | Check if the selected tag has the specified classe(s)

- Example:
    - Add class `"fade"` to tab panels, so that Bootstrap
    applies a *fade transition* between tabs:
    - Code:
    ```{r}
    tagQuery(tabs)$
      find(".tab-pane")$
      addClass("fade")$
      selectedTags()
    ```

### Altering element/children/siblings {.unnumbered}

| Method | Description |
|:---:|:---:| 
append | Insert content after the children of each selected tag
prepend | Insert content before the children of each selected tag
empty | Remove all children from the selected tag
remove | Remove all selected tags
before | Insert content before each selected tag
after | Insert content after each selected tag
replaceWith | Replace the currently selected tag by the provided tag

- Example:
    - For `tabs`, include an icon before each tab title 
    (text content of `<a>` element).
    - Code:
    ```{r}
    new_tabs <- tagQuery(tabs)$
      find("a")$
      prepend(icon("flag"))
    
    new_tabs$selectedTags()
    ```

### Chain tag queries {.unnumbered}

- The `$resetSelected()` function
allows us to **reset** the element selection
to the *base element* after a given operation,
making it possible to chain multiple queries.

- Let's combine all previous examples:
  ```{r}
  # Set HTML element "tabs" as base
  tagQuery(tabs)$
    # Operation 1:
    #   Add "fade" class to all panels
    find(".tab-pane")$
    addClass("fade")$
    resetSelected()$
    # New operation: 
    #   Add icon before the title of each "link"
    find("a")$
    prepend(icon("flag"))$
    allTags()
  ```

### Specific cases {.unnumbered}

- Some methods, like `$append()` and `$prepend()`
are only able to add the **same** elements
to one or multiple targets.

- When the transformation to perform is **index specific**,
`$each()` should be used. Such function takes an anonymous
function as input, with parameters **element** and **index**.

- Example:
    - Replace text content of each tab panel with
    **This is tab "tab_index"**.
    - Code:
    ```{r}
    tagQuery(tabs)$
      find(".tab-pane")$
      empty()$
      each(function(el, i) {
        el <- tagAppendChild(el, paste("This is tab", i))
        return(el)
      })$
      selectedTags()
    ```

### Practice {.unnumbered}

- Let's perform the same transformation
as in the **Practical examples** section,
but now using **tagQuery()**.

- Model:

    ```html
    <div>
      <label id="variable-label" for="variable">Variable:</label>
      <select id="variable" class="some-class">
        <option value="cyl" selected>Cylinders</option>
        <option value="gear">Gears</option>
      </select>
    </div>
    ```

- Example:
    ```{r}
    shinySelect <- shiny::selectInput(
      inputId = "selectId",
      label = "Choose",
      choices = 1:2,
      selected = 1
    )
    shinySelect
    ```

- Solution:
    ```{r}
    tagQuery(shinySelect)$
      # Remove class of container
      removeAttrs("class")$
      # Remove label element's class
      find(".control-label")$
      removeAttrs("class")$
      # Remove extra div container for <select>
      siblings()$
      replaceWith(
        tagQuery(shinySelect)$
          find(".control-label")$
          siblings()$
          children()$
          selectedTags()
      )$
      allTags()
    ```

### Alter tag rendering with rendering hooks {.unnumbered}

#### Simple hooks

- Let's explore one of the more **advanced** features of
the **htmltools** package: *Conditionally render an*
*HTML element, depending on conditions like external options*
*, a specific theme version, etc.*.

- As an example, the condition will be if Shiny is in 
`dev` mode:
    ```{r}
    devmode(TRUE)
    getOption("shiny.devmode")
    ```

- We create an htmltools tag with specific
instructions for the **.renderHook** parameter:
    ```{r}
    cssStyle <- "color: red; border: dashed red;"
    customTag <- span("", .renderHook = function(x) {
      if (getOption("shiny.devmode")) {
        return(tagAppendAttributes(x, style = cssStyle))
      }
    })

    customTag
    ```

- If the htmlools tag already has an existing **hook**,
the `tagAddRenderHook()` adds another hook to the current list:
    ```{r}
    customTag <- tagAddRenderHook(customTag, function(x) {
      if (getOption("shiny.devmode")) {
        return(tagAppendChildren(x, "UNDER REWORK"))
      } 
    })

    customTag$.renderHooks

    # Remove first hook
    # customTag$.renderHooks[[1]] <- NULL
    ```

- Let's try it in a Shiny app, after enabling
or disabling `dev` mode:
    ```{r, eval = FALSE}
    library(shiny)

    devmode(TRUE)
    # devmode(devmode = FALSE)
    getOption("shiny.devmode")

    cssStyle <- "color: red; border: dashed red;"
    customTag <- span("text", .renderHook = function(x) {
      if (getOption("shiny.devmode")) {
        return(tagAppendAttributes(x, style = cssStyle))
      }
    })

    ui <- fluidPage(
        customTag
      )
    server <- function(input, output) {}

    shinyApp(ui, server)
    ```

#### Nested hooks {.unnumbered}

- Most Shiny elements are composed of 
main wrappers and **nested** tags, for example,
`tabsetPanel()` and `tabPanel()`.

- How do we handle nested elements?

- Example:
```{r}
my_wrapper_tag <- function(...) {
  wrapper <- div(class = "parent", ...)
  items <- list(...)
  
  tagAddRenderHook(wrapper, function(x) {
    version <- getOption("theme_version")
    if (!is.null(version)) {
      if (version == "4") {
        new_items <- tagQuery(items)$
          find(".new-child")$
          each(function(x, i) {
            tagAppendAttributes(x, id = letters[i])
          })$
          allTags()
        
        x <- tagQuery(x)$
          empty()$
          append(new_items)$
          allTags()
      } 
    }
    return(x)
  })
}
```

- Let's design the nested item function:
```{r}
my_nested_tag <- function() {
  wrapper <- div(class = "nested")
  
  tagAddRenderHook(wrapper, function(x) {
    version <- getOption("theme_version")
    if (!is.null(version)) {
      x <- if (version == "4") {
        new_child <- div(class = "new-child")
        
        tagQuery(x)$
          append(new_child)$
          allTags()
      } 
    }
    return(x)
  })
}
```

- Testing:
```{r}
options("theme_version" = "4")
my_wrapper_tag(my_nested_tag(), my_nested_tag())

options("theme_version" = "3")
my_wrapper_tag(my_nested_tag(), my_nested_tag())
```

- Why did the `id` attribute did **not** get updated?
    - It's basically a **resolve issue**.
    - In the top-level render hook (`my_wrapper_tag`), the newly added item
    is not yet available. 
    - A **solution** is to resolve the sub items with `as.tags()`,
    in order to convert any arbitrary element to be a part of the tag structure.
    
```{r}
# Fixing the top-level render hook
my_wrapper_tag <- function(...) {
  wrapper <- div(class = "parent", ...)
  items <- list(...)
  
  tagAddRenderHook(wrapper, function(x) {
    version <- getOption("theme_version")
    if (!is.null(version)) {
      if (version == "4") {
        ###############################
        items <- lapply(items, as.tags)
        ###############################

        new_items <- tagQuery(items)$
          find(".new-child")$
          each(function(x, i) {
            tagAppendAttributes(x, id = letters[i])
          })$
          allTags()
        
        x <- tagQuery(x)$
          empty()$
          append(new_items)$
          allTags()
      } 
    }
    return(x)
  })
}
```

- Final test:

```{r}
options("theme_version" = "4")
my_wrapper_tag(my_nested_tag(), my_nested_tag())

options("theme_version" = "3")
my_wrapper_tag(my_nested_tag(), my_nested_tag())
```

A real-life case study is available later in the book in 
**section 9.3.3.3** .

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
